Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-05-27 paper/paper.tex
Final version
final post extension

This section provides a consistency proof for B2B in the case of low noise.

\begin{theorem}[Low-noise consistency]
    Consider the B2B model from Equation~\ref{eq:model}.
    %
    Assume that $F$ and $X$ are full-rank on the subspace spanned by active entries in $E$.
    %
    Then, as the variance of $N$ tends to zero and the sample size tends to infinity, B2B recovers $E$.
\end{theorem}
\begin{proof}
For centered noise $N$, the first regression satisfies:
\begin{align*}
    \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2].
\end{align*}
%
The solution to this first regression is $\hat{G} = [(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X$.

The second regression is:
%
\begin{align*}
    \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
\end{align*}
%
The solution to this second regression is $H = EF\hat{G}$.
%
Therefore, our estimate of $E$ is:
\begin{equation}
    \hat{E} = EF[(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X.
    \label{eq:estimate}
\end{equation}
%
By definition of the Moore-Penrose inverse (also known as pseudo-inverse), if $(NF)^\top (NF)$ is full-rank with $\|~(NF)^\top~(NF)~\|~\to~0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
%
Next, assume that the true causal influence matrix contains $k$ ones arranged as $E = \left(\begin{array}{cc} I_k & 0 \\ 0 & 0 \end{array}\right)$, and let $E = K^\top K$, where $K\in \mathbb{R}^{k \times d_x}$.
%
    Also, let the covariance matrix of $X$ be $X^\top X = \left(\begin{array}{cc}\Sigma_{1} & \Sigma_{2} \\ \Sigma_{3} & \Sigma_{4}\end{array}\right)$, where $\Sigma_1 \in \mathbb{R}^{k\times k}$.
%
Then,
\begin{align*}
    \hat{E} &= EF(XEF)^\dagger X = K^\top K F(XK^\top K F)^\dagger X \stackrel{i)}{=} K^\top (K F) (KF)^\dagger (XK^\top)^\dagger X\\
            &\stackrel{ii)}{=} K^\top K(XK^\top)^\dagger X = E (XK^\top)^\dagger X \stackrel{iii)}{=} E (KX^\top X K^\top)^{-1} KX^\top X\\
            &= E (KX^\top X K^\top)^{-1} (\Sigma_1 \,|\, \Sigma_2)  = E \Sigma_1^{-1} (\Sigma_1 \,|\, \Sigma_2)
            = \left(\begin{array}{cc} I_k & \Sigma_1^{-1} \Sigma_2 \\ 0 & 0 \end{array}\right),
\end{align*}
where the equalities follow due to i) a full-rank factorization of $XEF$, ii) $KF$ being full-rank yields $(KF)(KF)^\dagger = K$, and iii) $X K^\top$ being full-rank yields $(XK^\top)^\dagger = (KX^\top X K^\top)^{-1} KX^\top$.
%
Therefore, $\text{Diag}(\hat{E})$ is the desired causal influence matrix $E$.
\end{proof}

We observe that, even in the presence of strong noise, the expression of our estimate \eqref{eq:estimate} involves the left-most multiplication with the true causal influence matrix $E$.
%
This means that our estimate will, in expectation, contain zeros for all features that do not causally influence $Y$.
%
When the variance of $N$ increases, we of course pay a price: since the norm $\| X - XEF\hat{G} \| = \| X - X\hat{H} \|$ increases, this means that the values of $\hat{E}$ associated to the causal factors will be dampened.

@@

We perform two experiments to evaluate the performance of B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
%
We use scikit-learn to implement all of our simulations \citep{sklearn}.

>>>>

\begin{theorem}[B2B consistency - general case]

     Consider the B2B model from Equation $Y = (XE + N)F$, $N$ centred and full rank noise.
     %
     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat H$, will minimise
     %
     $\min_H  \left \| X - XH\right\| ^2  + \left \| NH\right \| ^2$and satisfy $E\hat H = \hat H$
\end{theorem}
\begin{proof}

 Let $\hat G$ and $\hat H$ be the solutions of the first and second regressions of B2B, we have
 \begin{align*}
    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2] &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2]\\
%                                                        &{}= \arg \min_G \mathbb{E}[\left \| X - XEFG\right\| ^2]  + \mathbb{E}[\left \| NFG\right \| ^2]\\
                                                        &{}= \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
     \label{eq:doublenorm}
\end{align*}
\begin{align*}
    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
%    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]\\
    &= EF \hat G
\end{align*}

Let us prove that $EF\hat G = F\hat G$, that is, that the $d_x$ bottom rows of $F\hat G$ are zero. Let $Z=F^\dagger EF\hat G$. We have $FZ = FF^\dagger EF  \hat G= EF\hat G$ ($FF^\dagger E =E$ since $F$ is full rank on $Img(E)$). Since E is a contraction, we have $ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$. Therefore, 
 $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$

But as $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, the above inequality is an equality, $Z=\hat G$ and $EF\hat G = F\Hat G$. Therefore, $\left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 = \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 = \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2$ is minimised by $\hat G$ and $\hat H$. Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$. This completes the proof.
\end{proof}

So, $\hat H = \arg \min_H  \left \| X - XEH\right\| ^2  + \left \| NEH\right \| ^2 = (E X^\top XE +EN^\top NE) ^\dagger EXX^\top$.

Assuming, without loss of generality, that the active features are the $k$ first, we have

$X^\top X = \left(\begin{array}{lccl}\Sigma_{X1} & \Sigma_{X2} \\ \Sigma_{X2} & \Sigma_{X3}\end{array}\right)$, $N^\top N = \left(\begin{array}{lccl}\Sigma_{N1} & \Sigma_{N2} \\ \Sigma_{N2} & \Sigma_{N3}\end{array}\right)$ and

$\hat H = \left(\begin{array}{cc}(\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1} & (\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X2} \\0 & 0\end{array}\right)$

$Diag_k (\hat H) = Diag((\Sigma_{X1}+\Sigma_{N1})^{-1}\Sigma_{X1}) = Diag((I+\Sigma_{X1}^{-1}\Sigma_{N1})^{-1})$

In the absence of noise, we have $Diag(\hat H) = Diag(E)$. Else the k first diagonal elements of $\hat H$ are all positive, and bounded by $\frac{\sigma_{X_k}}{\sigma_{X_k} +\sigma_{N_1}}$ and $\frac{\sigma_{X_1}}{\sigma_{X_1} +\sigma_{N_k}}$, where  $\sigma_{X_1}$, $\sigma_{X_k}$, $\sigma_{N_1}$ and $\sigma_{N_k}$ denote the largest and smallest eigenvalues of $\Sigma_{X1}$ and $\Sigma_{N1}$.

The average value of the coefficients of $Diag(\hat H)$ is the trace of $\hat H$ divided by k. On average, since $X$ and $N$ are not correlated, $\Sigma_{X1}^{-1}\Sigma_{N1}$ should have a low condition numbers, and $\frac{Var(X)}{Var(X)+Var(N)}$ provide a good estimate of $\frac{1}{k}Tr(\hat H)$.

@@

We perform two experiments to evaluate the performance of B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
%
We use scikit-learn \citep{sklearn}.
